How do we copy and paste? The semantic drift of quotations in blogspace

### 2015-03-06 brains-conclusion.tex
Phrasing fixes, blue text to validate

For each of the selected psycholinguistic variables, we demonstrate the existence of attractor values in the underlying variable spaces. More precisely, beyond the interpretation of our results for each variable, we notice that all variables remarkably exhibit a single attractor and are generally contractile --- as such, even though the observed convergence patterns only partially explain quotation evolution, we shed light on a class of phenomena which are prone to constitute a key element of a broader empirically-grounded, attractor-based theory of cultural evolution.

>>>>

For each of the selected psycholinguistic variables, we demonstrate the existence of attractor values in the underlying variable spaces. More precisely, beyond the interpretation of our results for each variable, we notice that all variables remarkably exhibit a single attractor and are generally contractile --- as such, even though the observed convergence patterns only partially explain quotation evolution, we shed light on a class of phenomena which are susceptible to constitute a key element of a broader empirically-grounded, attractor-based theory of cultural evolution.

### 2015-03-06 brains-introduction.tex
Phrasing fixes, blue text to validate

Broadly speaking, we may distinguish two main research streams, depending on whether the focus lies on cognitive processes or on social dynamics. A first stream is largely structured around cultural anthropology and essentially addresses cultural similarity, diversity and its evolution. It features several theories mixing social and individual cognition including, to cite a few, the debated ``memetic'' program initiated by \citet{Dawkins76} (for which the collection of works by \citealp{Aunger00}, provides a solid overview), the development of evolutionary models of norms~\citep[see for instance][]{Ehrlich05} following the seminal work of \citet{Boyd85}; or the ``cultural epidemiology'' program proposed by \citet{sper:expl}, which links the concept of mental representation to the concept of public representation (the latter being the counterpart of the former outside of the brain, \hbox{i.e.} in all sorts of cultural artifacts: texts, utterances, etc.).
%\footnote{\citet{sper:expl} emphasizes this distinction in his seminal work: ``A representation may exist inside its user: it is then a \emph{mental representation}, such as a memory, a belief, or an intention. The producer and the user of a mental representation are one and the same person. A representation may also exist in the environment of its user, as is the case, for instance, of the text you are presently reading: it is then a \emph{public representation}''.}

@@

At this level, we have to turn to the broader psycholinguistic literature which provides one of the main cognitive foundations for public representation evolution by studying the influence of word features on the ease of recall.
This field is well developed and details the impact that classical psycholinguistic variables such as word frequency~\citep[see][for a review]{Yonelinas02}, age-of-acquisition~\citep{Zevin02}, number of phonemes or number of syllables~\citep[see for instance][]{Rey98,nick-diss}, have in this type of task.

@@

On the whole, the current psycholinguistic state-of-the-art seems to hint towards two antagonistic types of results.
On one hand, part of the literature tends to show that recall is easier for the least ``awkward'' words; those whose age of acquisition is earlier, length is smaller, semantic network position is more central -- this is particularly true in tasks where participants are asked to form spontaneous associations or utter a word in response to a given signal.
On the other hand, when the task consists in recognizing a specific item in a list, ``awkward'' words are actually more easily remembered, possibly as they are more informative and plausibly more discernible~\citep[see again][for a review]{Yonelinas02}.
The jury is still out as to whether reformulation alteration, that is spontaneous replacement of words when asked to repeat a given utterance, is rather of the former or latter sort.
We also aim here at shedding some light on this debate, considering oddness as a dimension of the purported fitness of utterances.

>>>>

Broadly speaking, we may distinguish two main research streams, depending on whether the focus lies on cognitive processes or on social dynamics. A first stream is largely structured around cultural anthropology and essentially addresses cultural similarity, diversity and its evolution. It features several theories mixing social and individual cognition including, to cite a few, the debated ``memetic'' program initiated by \citet{Dawkins76} (for which the collection of works by \citealp{Aunger00}, provides a solid overview), the development of evolutionary models of norms~\citep[see for instance][]{Ehrlich05} following the seminal work of \citet{Boyd85}; or the ``cultural epidemiology'' program proposed by \citet{sper:expl}, which links the concept of mental representation to the concept of public representation (the latter being the counterpart of the former outside of the brain, \hbox{i.e.} in all kinds of cultural artifacts: texts, utterances, etc.).
%\footnote{\citet{sper:expl} emphasizes this distinction in his seminal work: ``A representation may exist inside its user: it is then a \emph{mental representation}, such as a memory, a belief, or an intention. The producer and the user of a mental representation are one and the same person. A representation may also exist in the environment of its user, as is the case, for instance, of the text you are presently reading: it is then a \emph{public representation}''.}

@@

At this level, we have to turn to the broader psycholinguistic literature which provides one of the main cognitive foundations for public representation evolution by studying the influence of word features on the ease of recall.
This field is well developed and details the impact that classical psycholinguistic variables such as word frequency~\citep[see][for a review]{Yonelinas02}, age-of-acquisition~\citep{Zevin02}, number of phonemes or number of syllables~\citep[see for instance][]{Rey98,nick-diss}, have on this type of task.

@@

On the whole, the current psycholinguistic state-of-the-art seems to hint towards two antagonistic types of results.
On one hand, part of the literature tends to show that recall is easier for the least ``awkward'' words; those whose age of acquisition is earlier, length is smaller, semantic network position is more central --- this is particularly true in tasks where participants are asked to form spontaneous associations or utter a word in response to a given signal.
On the other hand, when the task consists in recognizing a specific item in a list, ``awkward'' words are actually more easily remembered, possibly as they are more informative and plausibly more discernible~\citep[see again][for a review]{Yonelinas02}.
The jury is still out as to whether reformulation alteration, that is spontaneous replacement of words when asked to repeat a given utterance, is rather of the former or latter sort.
We also aim here at shedding some light on this debate, considering oddness as a dimension of the purported fitness of utterances.

### 2015-03-06 brains-protocol.tex
Phrasing fixes, blue text to validate

We could therefore examine the individual substitution process at work when authors transform quotations, by examining the features of the substituted and substituting words in each substitution. To keep the analysis palatable, we focused on quotation transformations consisting in the \emph{substitution} of a word by another word (and only those cases) in order to unambiguously discuss single word replacements. 
To quantify those substitutions, we decided to associate a number of features to each word, the variation of which we can statistically study.

@@

We used a quotation dataset collected by \citet{Leskovec09}, large enough to lend itself to statistical analysis.
This dataset consists of the daily crawling of news stories and blog posts from around a million online sources, with an approximate publication rate of 900k texts per day, over a nine-month period of time (from August 2008 to April 2009~---~\citealp{Leskovec09-url}).\footnote{Unfortunately, the original article~\citep{Leskovec09} does not provide additional details on the source selection methodology.}
Quotations were then automatically extracted from this corpus: each quotation is a more or less faithful excerpt of an utterance (oral or written) by the quoted person. For instance,
\begin{quote}
The Bank of England said, ``these operations are designed to address funding pressures over quarter-end.''
\end{quote}
Quotations were then gathered in a graph and connected according to their similarity: either because they differ by very few words (in that case, no more than one word) or because they share a certain sequence of words (in that case, at least ten consecutive words). We find for example the following variation of the above quote: \begin{quote}``these operations are \textbf{intended} to address funding pressures over quarter-end.''\end{quote}
A community detection algorithm was applied to that quotation graph to detect aggregates of tightly connected, that is sufficiently similar, groups of quotations~\citep[see][for more detail]{Leskovec09}.
This analysis yielded the final data we had access to, with a total of about \num{70000} sets of quotations; each of these sets allegedly contains all variations of a same parent utterance, along with their respective publication URLs and timestamps.

@@

Note that since we only consider substitutions and not faithful copies, we measure the features of an alteration \emph{knowing that there has been an alteration}, and we do not take invariant quotations into account.
Indeed, in the former case we know there has been a human reformulation, whereas in the latter case it is impossible to know whether there has been perfect human reformulation or simply digital copy-pasting of a source (``{\sc Ctrl-C}/{\sc Ctrl-V}'').
Furthermore, perfect human reformulation possibly involves different practices than those involved in alteration ---~for instance drafting before publishing, double-checking sources, proof-reading~--- and may not be representative of the cognitive processes at work during alteration.
The two situations are different enough to be studied separately, and we focus here on the latter.

@@

\begin{figure*}[!th]
    \centering
    \includegraphics[width=\textwidth]{images/computed-figures/feature_susceptibilities.png}
    \caption{\textbf{Substitution susceptibility:} average susceptibility to substitution versus average feature value of a candidate word for substitution, with 95\% asymptotic confidence intervals.
    The heatmap on the lower-right shows the joint effect of Number of synonyms and Number of phonemes on susceptibility, averaged over the respective single-variable ranges, with sample size (word numbers) in parentheses.}
    \label{fig:feature-susceptibilities}
\end{figure*}
\medskip
Results for this measure are gathered in Fig.~\ref{fig:feature-susceptibilities}. They first show an obvious strong effect of Word frequency: the more frequent a word, the less likely it is to attract substitutions.
Indeed, susceptibility goes from $.33$ for low-frequency words down to nearly 0 for very high-frequency words.
To make things clear, this value of $.33$ means that low-frequency words, when present in a quote undergoing a substitution, are the ones being substituted 33\% of the time on average.

The other features --- Age of acquisition, Number of phonemes, Clustering coefficient and Number of synonyms --- do not seem to exhibit any particularly significant effect on susceptibility.
{If we set aside the values for low Number of phonemes}, for each of these features it is indeed possible to draw a constant line which always remains within the respective confidence intervals.
If these variables have an effect, it is by no means as strong as it is for Word frequency.
This is remarkably clear for Clustering coefficient and Age of acquisition, where susceptibility values remain within quite small intervals (respectively $[.13-.18]$ and $[.16-.20]$).
{We may notice a slight effect for the lowest values of Number of synonyms and Number of phonemes, where the mean susceptibility is almost half as high as the average of the other values (respectively $.09$ vs. $.16$, and $.11$ vs. $.17$).}
Keeping in mind the poor statistical significance of this effect, we could still wonder if the shortest words and words with fewest synonyms are significantly less susceptible to substitution.
To further examine this phenomenon, we plotted the two-dimensional map of susceptibility values for these two features (see heatmap at the bottom right of Fig.~\ref{fig:feature-susceptibilities}).
%\TG{TBF: [[We see that outlier values (the two darkest cells) are both due to cells where the number of items is only slightly above the significance threshold (11 items, while we decided to cut at 10).]]}
%\TG{TBF: [[If we focus on cells based on at least several dozens of words]]} 
Values tend to navigate around the mean value ($.16$), even if there are a few outlier cells, yet with little obvious regularity (except for a low number of synonyms, consistently with the unidimensional graph). {On the whole, this makes it relatively hard to draw any conclusion as regards the direction of an effect, except for the least populated value ranges (which as a result are also less significant).}
%\TG{TBF: [[This confirms the fact that the slight late increases observed for Age of acquisition and Number of phonemes are rather non-significant.]]}

@@

All in all, apart from Word frequency and despite some local tendencies, {in general} these results do not allow us to conclude either to a marked effect %or to no effect 
of the selected psycholinguistic features on substitution susceptibility.
We may therefore globally assume that substitution targets are chosen in a more or less uniform way with respect to these features.

@@

We can thus show how words are modified once we know they are substituted, that is how their features are modified by said substitution.
Considering a word $\wstart$ substituted for $\warrival$, we measure how the feature of $\wstart$ varies when it is replaced with $\warrival$, that is we look at $\phi(\warrival)$ as a function of $\phi(\wstart)$. 
Averaging this value over all start words such that $\phi(w) = f$ yields the mean variation for that feature value~$f$, that is:\footnote{To avoid possible autocorrelation effects due to substitutions belonging to the same cluster (which are likely not statistically independent and may lead to overly optimistic confidence intervals), we first average substitutions over each cluster, by considering the average of arrival word features for a given start word.}
$$\nu_{\phi}(f) = \left<\phi(\warrival)\right>_{\left\lbrace \wstart\rightarrow\warrival | \phi(\wstart) = f \right\rbrace}$$

>>>>

\TB{We could therefore study the individual transformation process at work when authors alter quotations, by examining the modified words in each transformation.} To keep the analysis palatable, we focused on quotation transformations consisting in the \emph{substitution} of a word by another word (and only those cases) in order to unambiguously discuss single word replacements.
To quantify those substitutions, we decided to associate a number of features to each word, the variation of which we can statistically study.

@@

We used a quotation dataset collected by \citet{Leskovec09}, large enough to lend itself to statistical analysis.
This dataset consists of the daily crawling of news stories and blog posts from around a million online sources, with an approximate publication rate of 900k texts per day, over a nine-month period of time (from August 2008 to April 2009~---~\citealp{Leskovec09-url}).\footnote{Unfortunately, the original article~\citep{Leskovec09} does not provide additional details on the source selection methodology.}
Quotations were then automatically extracted from this corpus: each quotation is a more or less faithful excerpt of an utterance (oral or written) by the quoted person. For instance,
\begin{quote}
The Bank of England said, ``these operations are designed to address funding pressures over quarter-end.''
\end{quote}
Quotations were then gathered in a graph and connected according to their similarity: either because they differ by very few words (in that case, no more than one word) or because they share a certain sequence of words (in that case, at least ten consecutive words). We find for example the following variation of the above quote: \begin{quote}``these operations are \textbf{intended} to address funding pressures over quarter-end.''\end{quote}
A community detection algorithm was applied to that quotation graph to detect aggregates of tightly connected, that is sufficiently similar, groups of quotations~\citep[see][for more details]{Leskovec09}.
This analysis yielded the final data we had access to, with a total of about \num{70000} sets of quotations; each of these sets allegedly contains all variations of a same parent utterance, along with their respective publication URLs and timestamps.

@@

Note that since we only consider substitutions and not faithful copies, we measure the features of an alteration \emph{knowing that there has been an alteration}, and we do not take invariant quotations into account.
Indeed, in the former case we know there has been a human reformulation, whereas in the latter case it is impossible to know whether there has been perfect human reformulation or simply digital copy-pasting of a source~(``{\sc Ctrl-C}/{\sc Ctrl-V}'').
Furthermore, perfect human reformulation possibly involves different practices than those involved in alteration ---~for instance drafting before publishing, double-checking sources, proof-reading~--- and may not be representative of the cognitive processes at work during alteration.
The two situations are different enough to be studied separately, and we focus here on the latter.

@@

\begin{figure*}[!th]
    \centering
    \includegraphics[width=\textwidth]{images/computed-figures/feature_susceptibilities.png}
    \caption{\textbf{Substitution susceptibility:} average susceptibility to substitution versus average feature value of a candidate word for substitution, with 95\% asymptotic confidence intervals.
    The heatmap on the lower-right shows the joint effect of Number of synonyms and Number of phonemes on susceptibility, averaged over the respective single-variable ranges, with sample size (word numbers) in parentheses.}
    \label{fig:feature-susceptibilities}
\end{figure*}
\medskip
Results for this measure are gathered in Fig.~\ref{fig:feature-susceptibilities}. They first show an obvious strong effect of Word frequency: the more frequent a word, the less likely it is to attract substitutions.
Indeed, susceptibility goes from $.33$ for low-frequency words down to nearly~0 for very high-frequency words.
To make things clear, this value of $.33$ means that low-frequency words, when present in a quote undergoing a substitution, are the ones being substituted 33\% of the time on average.

The other features --- Age of acquisition, Number of phonemes, Clustering coefficient and Number of synonyms --- do not seem to exhibit any particularly significant effect on susceptibility.
{If we set aside the values for low Number of phonemes}, for each of these features it is indeed possible to draw a constant line which always remains within the respective confidence intervals.
If these variables have an effect, it is by no means as strong as it is for Word frequency.
This is remarkably clear for Clustering coefficient and Age of acquisition, where susceptibility values remain within quite small intervals (respectively $[.13-.18]$ and $[.16-.20]$).
{We may notice a slight effect for the lowest values of Number of synonyms and Number of phonemes, where the mean susceptibility is almost half as high as the average of the other values (respectively $.09$ vs. $.16$, and $.11$ vs. $.17$).}
Keeping in mind the poor statistical significance of this effect, we could still wonder if the shortest words and words with fewest synonyms are significantly less susceptible to substitution.
To further examine this phenomenon, we plotted the two-dimensional map of susceptibility values for these two features (see heatmap at the bottom right of Fig.~\ref{fig:feature-susceptibilities}).
%\TG{TBF: [[We see that outlier values (the two darkest cells) are both due to cells where the number of items is only slightly above the significance threshold (11 items, while we decided to cut at 10).]]}
%\TG{TBF: [[If we focus on cells based on at least several dozens of words]]}
\TB{Even if there are a few outlier cells, values tend to navigate around the mean value ($.16$) with little obvious regularity (except for a low number of synonyms, consistent with the unidimensional graph)}. {On the whole, this makes it relatively hard to draw any conclusion as regards the direction of an effect, except for the least populated value ranges (which as a result are also less significant).}
%\TG{TBF: [[This confirms the fact that the slight late increases observed for Age of acquisition and Number of phonemes are rather non-significant.]]}

@@

All in all, apart from Word frequency and despite some local tendencies, {in general} these results do not allow us to conclude \TB{\sout{either}} to a marked effect %or to no effect
of the selected psycholinguistic features on substitution susceptibility.
We may therefore globally assume that substitution targets are chosen in a more or less uniform way with respect to these features.

@@

We can thus show how words are modified once we know they are substituted, that is how their features are modified by said substitution.
Considering a word~$\wstart$ substituted for~$\warrival$, we measure how the feature of~$\wstart$ varies when it is replaced with~$\warrival$, that is we look at~$\phi(\warrival)$ as a function of~$\phi(\wstart)$.
Averaging this value over all start words such that $\phi(w) = f$ yields the mean variation for that feature value~$f$, that is:\footnote{To avoid possible autocorrelation effects due to substitutions belonging to the same cluster (which are likely not statistically independent and may lead to overly optimistic confidence intervals), we first average substitutions over each cluster, by considering the average of arrival word features for a given start word.}
$$\nu_{\phi}(f) = \left<\phi(\warrival)\right>_{\left\lbrace \wstart\rightarrow\warrival | \phi(\wstart) = f \right\rbrace}$$
